---
title: "R ladies Ottawa Meeting June 2019"
author: "Deirdre Hennessy"
date: "June 12th, 2019"
output:
  prettydoc::html_pretty:
    highlight: github
    theme: cayman
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
library(DataExplorer)
library(ggplot2)
library(readr)

```

## Data manipulation with dplyr and the iris dataset

The tidyverse, which includes dplyr, is a collection of packages used for reading data, data manipulation, graphing and other things. 

https://www.tidyverse.org/packages/

https://r4ds.had.co.nz/tidy-data.html


"Stunning is one of the most useful spells in your arsenal. It's sort of a wizard's bread and butter, really." -Harry Potter regarding the usefulness of the spell.

"The tidyverse is one of the most useful collections of packages in your arsenal. It's sort of a analyst's bread and butter, really." -Deirdre Hennessy  regarding the usefulness of the tidyverse. 




## Look at my data

First I am going to take a look at my data.


```{r}

head(iris)
str(iris)
summary(iris)

# If I wanted to get a little fancier I could use the DataExplorer package to create a report

create_report(iris)
```

# 1. Use select to reorder and rename to ..........rename (select)


Next I am going to arrange the data set the way I want to see it.

```{r}

iris_df<-iris %>% 
  select(Species:Sepal.Length)  %>%  # to reorder
  
  rename( petal_width = Petal.Width)  %>% # to rename
  rename( petal_length = Petal.Length) %>% 
  rename( sepal_width = Sepal.Width)  %>% 
  rename( sepal_length = Sepal.Length)  

head(iris_df)

```

# 2. Select the columns I require (select)

```{r}

iris_df_sepal<- iris_df  %>% 
  select(species=Species, sepal_width, sepal_length)

head(iris_df_sepal)

# I want to look at the distribution of sepal length

summary(iris_df_sepal$sepal_length)

ggplot(iris_df_sepal, aes(sepal_length)) +
  geom_histogram(bins = 25)


```

# 3. It might be interesting to look at sepal length of greater than 5- filter the rows I require (filter)

```{r}

iris_df_sepal<- iris_df  %>% 
  select(species=Species, sepal_width, sepal_length)   %>% 
  filter(sepal_length>=5.0) %>% 
  filter(species== "virginica") 

# or it is possible to add mulitple filter conditions to the same line 

iris_df_sepal<- iris_df  %>% 
  select(species=Species, sepal_width, sepal_length)   %>% 
  filter(sepal_length>=5.0, species== "virginica")

head(iris_df_sepal)


```

# 4. Now I want to transform my variables for analysis- create a new ratio varaible (mutate)

```{r}

iris_df_sepal<- iris_df  %>% 
  select(species=Species, sepal_width, sepal_length)   %>% 
  filter(sepal_length>=5.0, species== "virginica")  %>%
  
  mutate(wlratio=sepal_width/sepal_length) #

head(iris_df_sepal)
```

# 5. Next I want to summarise my variables by species (group_by and summarise )

```{r}

iris_df_all<- iris_df  %>% 
  select(species=Species, sepal_width, sepal_length, petal_width, petal_length)   %>% 
  mutate(wlratio_sepal=sepal_width/sepal_length)  %>%
  mutate(wlratio_petal=petal_width/petal_length)  %>%
   
  group_by(species) %>%
  summarise(mean_sepalw=mean(sepal_width), mean_sepall=mean(sepal_length),
            mean_petalw=mean(petal_width), mean_petall=mean(petal_length),
            meanwl_sep=mean(wlratio_sepal), meanwl_pet=mean(wlratio_petal))

head(iris_df_all)

```

# 6. Playing with spread and gather.....we need more complex data

```{r}

# Data on projected diabetes cases

diabetes <- read.csv("C:/Users/colli/Desktop/PHAC_Infobase_CCDSS.csv")

# Select and filter what I need

head(diabetes)

diabetes_red <- diabetes %>%
  select(Sex, Year, province=Area, prevalence=Rate, Age= Age.Group)  %>%
  filter(province == 'Canada', Sex == 'Both sexes', Year > 2000, Age == "'20-34'") %>%
  
  spread(key=Year, value=prevalence) 

head(diabetes_red)

diabetes_red <- diabetes %>%
  select(Sex, Year, province=Area, prevalence=Rate, Age= Age.Group)  %>%
  filter(province == 'Canada', Sex == 'Both sexes', Year > 2000, Age == "'20-34'") %>%
  
  spread(key=Year, value=prevalence) %>%
  
  gather('2001', '2002', '2003', '2004', '2005', '2006', '2007', '2008', '2009', '2010','2011', '2012', '2013', '2014', '2015', key = "Year", value = "proportion")

head(diabetes_red)
```


